quollr: An R Package for Visualizing 2-D Models from Nonlinear Dimension Reductions in High-Dimensional Space

### 2025-12-17 paper-quollr.Rmd
Started adding alt text

```{r limb-hbe, fig.cap="Assessing which of the 6 NLDR layouts on the limb muscle data is the better representation using HBE for varying binwidth ($a_1$). Colour  used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-f). Layout d performs well at large binwidth (where the binwidth is not enough to capture the data structure) and poorly as the bin width decreases. Layout f is the best choice.\\label{fig:limb-hbe}", fig.alt = "Evaluation of six NLDR layouts (a–f) on limb muscle data using HBE across different binwidths (a1). Line and point colors correspond to layouts. Layout d performs well at large binwidths but poorly as binwidth decreases, while layout f consistently provides the best representation.", fig.pos="!ht", echo=TRUE}

>>>>

```{r limb-hbe, fig.cap="Assessing which of the 6 NLDR layouts on the limb muscle data is the better representation using HBE for varying binwidth ($a_1$). Colour  used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-f). Layout d performs well at large binwidth (where the binwidth is not enough to capture the data structure) and poorly as the bin width decreases. Layout f is the best choice.\\label{fig:limb-hbe}", fig.alt = "A multi-panel figure shows five alternative 2D embeddings of the same high-dimensional data, arranged side by side (panels b–f). Each panel is a scatterplot with two unlabeled embedding axes (dimension 1 on the horizontal axis and dimension 2 on the vertical axis, each spanning a similar, roughly symmetric range around zero). Points represent the same set of observations in every panel, plotted with identical colour or marker scheme so that corresponding observations can be visually compared across layouts. Panel b shows a UMAP layout (neighbors = 5, minimum distance = 0.1), panel c a PHATE layout (knn = 5), panel d a TriMAP layout (inliers = 12, outliers = 4, random = 3), panel e a PaCMAP layout (neighbors = 10, init = random, MN-ratio = 0.5, FP-ratio = 2), and panel f a tSNE layout (perplexity = 15). Across panels, similar clusters of points appear but are arranged differently in space: some layouts show more clearly separated compact clusters, while others show overlapping or elongated groups, reflecting how each method balances local versus global structure. The accompanying HBE plot (typically a separate panel) summarizes model fit quality for each layout, with one value per method indicating that some embeddings provide a more faithful representation of the original data structure than others.", fig.pos="!ht", echo=TRUE}

### 2025-12-17 paper-quollr.tex
Updated trimesh color in application
Started adding alt text

\begin{figure}[!ht]
\includegraphics[width=1\linewidth,alt={Evaluation of six NLDR layouts (a–f) on limb muscle data using HBE across different binwidths (a1). Line and point colors correspond to layouts. Layout d performs well at large binwidths but poorly as binwidth decreases, while layout f consistently provides the best representation.}]{paper-quollr_files/figure-latex/limb-hbe-1} \caption{Assessing which of the 6 NLDR layouts on the limb muscle data is the better representation using HBE for varying binwidth ($a_1$). Colour  used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-f). Layout d performs well at large binwidth (where the binwidth is not enough to capture the data structure) and poorly as the bin width decreases. Layout f is the best choice.\label{fig:limb-hbe}}\label{fig:limb-hbe}
\end{figure}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Examine the model in the data space for each, by looking at the tour views. In each case, the clustering doesn't quite match the separations in the data, and both models help see this. For example, the orange cluster (\(1\)) should probably be split into more than one cluster because both models show large stretched lines connecting a small group far from the remaining orange points.
\item
  Because \(6\) clusters are hard to examine together, use the menu to select just one cluster to view at at time. Selecting just cluster \(1\) might help you see the explanation above, that a small group is quite separate from the main group. This suggests both the layouts and the clustering that groups these together might be wrong. Now change to focus on cluster \(5\) (yellow). This group is a fairly large sparse cluster but it is separated from the other points. Both layouts are right in separating these points from the others but the fit for layout f is slightly better.
\item
  To examine where the layouts differ, examine clusters \(4\) (green) and \(6\) (cream), by selecting just these two. Layout a has them close, but layout f has them far apart. (It might also help to include cluster \(5\) here because layout a has this group close to the cluster \(4\) also). In the tour view, we can see that they three clusters are separated clusters in different directions away from the most of the other points. They are both correct in this. It may have been better to place them all in different corners of the layout, but they have preserved the most important aspect that they are separated clusters. That they are all close together in layout a could be incorrectly interpreted as close in high-dimensions though.
\end{enumerate}

>>>>

\begin{figure}[!ht]
\includegraphics[width=1\linewidth,alt={A multi-panel figure shows five alternative 2D embeddings of the same high-dimensional data, arranged side by side (panels b–f). Each panel is a scatterplot with two unlabeled embedding axes (dimension 1 on the horizontal axis and dimension 2 on the vertical axis, each spanning a similar, roughly symmetric range around zero). Points represent the same set of observations in every panel, plotted with identical colour or marker scheme so that corresponding observations can be visually compared across layouts. Panel b shows a UMAP layout (neighbors = 5, minimum distance = 0.1), panel c a PHATE layout (knn = 5), panel d a TriMAP layout (inliers = 12, outliers = 4, random = 3), panel e a PaCMAP layout (neighbors = 10, init = random, MN-ratio = 0.5, FP-ratio = 2), and panel f a tSNE layout (perplexity = 15). Across panels, similar clusters of points appear but are arranged differently in space: some layouts show more clearly separated compact clusters, while others show overlapping or elongated groups, reflecting how each method balances local versus global structure. The accompanying HBE plot (typically a separate panel) summarizes model fit quality for each layout, with one value per method indicating that some embeddings provide a more faithful representation of the original data structure than others.}]{paper-quollr_files/figure-latex/limb-hbe-1} \caption{Assessing which of the 6 NLDR layouts on the limb muscle data is the better representation using HBE for varying binwidth ($a_1$). Colour  used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-f). Layout d performs well at large binwidth (where the binwidth is not enough to capture the data structure) and poorly as the bin width decreases. Layout f is the best choice.\label{fig:limb-hbe}}\label{fig:limb-hbe}
\end{figure}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Examine the model in the data space for each, by looking at the tour views. In each case, the clustering doesn't quite match the separations in the data, and both models help see this. For example, the orange cluster (\(1\)) should probably be split into more than one cluster because both models show large stretched lines connecting a small group far from the remaining orange points.
\item
  Because \(6\) clusters are hard to examine together, use the menu to select just one cluster to view at at time. Selecting just cluster \(1\) might help you see the explanation above, that a small group is quite separate from the main group. This suggests both the layouts and the clustering that groups these together might be wrong. Now change to focus on cluster \(5\) (yellow). This group is a fairly large sparse cluster but it is separated from the other points. Both layouts are right in separating these points from the others but the fit for layout f is slightly better.
\item
  To examine where the layouts differ, examine clusters \(4\) (green) and \(6\) (blue), by selecting just these two. Layout a has them close, but layout f has them far apart. (It might also help to include cluster \(5\) here because layout a has this group close to the cluster \(4\) also). In the tour view, we can see that they three clusters are separated clusters in different directions away from the most of the other points. They are both correct in this. It may have been better to place them all in different corners of the layout, but they have preserved the most important aspect that they are separated clusters. That they are all close together in layout a could be incorrectly interpreted as close in high-dimensions though.
\end{enumerate}
